# Usage

Postponing the definition of new models through model files to Chapter \@ref(the-model-file), the typical workflow when using **pdphmc** consists of the following three steps:

- Building the model.
- Fitting the model to a given data set.
- Using output of the fitting process.

Throughout this chapter, we consider the very simple model $y_i \sim N(\mu,1),\;i=1,\dots,n$ where $\{y_i\}_{i=1}^n$ are observed,
the single parameter $\mu$ has a flat prior, and we wish to draw samples
from the posterior $p(\mu|\{y_i\}_{i=1}^n)$.

(Okay, we know the posterior in closed form, $\mu|\{y_i\}_{i=1}^n \sim N(\bar y,n^{-1}),\; \bar y = n^{-1}\sum_{i=1}^n y_i$, but for the sake of illustration...)

A model file for this simple model, where we are for some reason also interested 
in the quantity $\mu^2$ may be something like the following:
```
INCLUDE_BLOCK{
  // not used here
}
DATA_BLOCK{
  DATA_VECTOR(y); // data to be passed from R
}
SETUP_BLOCK{
  // not used here
}
VARIABLES_BLOCK{
  PARAMETER_SCALAR(mu); // the parameter
  GENERATED_SCALAR(mu_squared); // a transformation of the parameter
}
MODEL_BLOCK{
  // log-likelihood function
  target += stan::math::normal_lpdf(y,mu,1); // note stan math library available
  // generated quantities
  mu_squared = mu*mu;
}
```
From now on, we assume the above code is available in a file titled 
**iidmodel.cpp** in the current working directory.


## Building a model
The first step for performing analysis is to "build" the model. This process involves two step:

- Translating the model file into a working c++ program (the code in the "blocks" should in all cases be valid c++ code)
- Compiling the program to obtain a stand alone binary that communicates with **R** using text files. 

Unless your model file contains error, this process should be completely hidden for the user. 

Building of the model is done using the function **build()**, e.g.

```{r eval=FALSE}
model<-pdphmc::build("iidmodel.cpp")
```

This process should result in a folder "iidmodel_files" containing a single standalone program. The S4 object ```model``` contains different metadata regarding the model and the path to the program generated.

If the model is not correctly specified, ```pdphmc::build()``` outputs the compiler errors generated, which may be used to diagonse the problem.

See the documentation for further arguments to ```pdphmc::build()```, in particular the event rate specification and the mass matrix are specified at this
stage.

## Running a model

The next step is to run the model using the function ```pdphmc::run()``` for a given data set. Here we simulate a data
set with $n=50$, and pass it to ```pdphmc::run()``` in a named list (name correspond to name given in the ```DATA_VECTOR()```-macro in the model file above).

```{r eval=FALSE}
set.seed(123)
data <- rnorm(50,mean=3.0) # the data set to be considered
fit <- run(model,data=list(y=data)) # output is stored in S4 object "fit"
```

## Handling output

Finally the output from ```pdphmc::run()``` may be used in different ways. Most importantly

* get summary statistics (based on ```rstan::monitor()```):
```{r eval=FALSE}
pdphmc::getMonitor(fit) 
#compare mean of mu against:
mean(data)
#compare sd of mu against:
sqrt(1.0/50.0)
```
* get trace plots to check convergence (trace plot also include warmup period)
```{r eval=FALSE}
pdphmc::trace.plot(fit,c("mu","mu_squared"))
```


**pdphmc** actually both records two types of samples from the underlying PDMP; discrete samples (which were displayed above, and may be used as regular MCMC samples) and continuous/integrated samples. The continuous samples **only have the correct first order moments** and should not be used for estimating standard deviations, quantiles etc. Continuous samples are only recorded for the GENERATED-quantities. The continuous samples may be extracted using e.g.

```{r eval=FALSE}
pdphmc::getIntMonitor(fit)
pdphmc::trace.plot.int(fit)
```
